---
title: "How to use the `tinyForestR` package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use the `tinyForestR` package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE, 
  cache = FALSE, 
  comment = "#>", 
  fig.width = 8, 
  fig.height = 6
)




```

## What is the `tinyForestR` package?

`tinyForestR` is a set of tools designed to extract, manipulate and analyse data relevant to the location of Tiny Forests in the UK.

Specifically it extracts and processes landcover and biodiversity data from a range of sources for a given area around Tiny Forest locations, and provides a set of tools for analysing citizen science data derived directly from Tiny Forests.

## Getting started

The package is hosted on Github and is a work in progress. It can be installed by running `devtools::install_github("julianflowers/tinyForestR")`.

The package makes use of a number of Application Programming Interfaces (APIs) some of which require API keys which will need to be applied for separately. This is outlined in the relevant sections of this vignette.

It also uses a range of Python packages to access some datasets (in some cases Python packages are better developed than R). For this reason the first step is to run `initialise_tf()` to intialise the package.

This:

-   Loads `reticulate` which R uses to talk to Python

-   Sets up a Python virtual environment (VE) called `tinyforest`, and ensures R uses the correct version of Python by specifying the `RETICULATE_PYTHON` environment variable. It may be necessary to restart R to make sure R correctly uses this.

-   You will be asked if you want to remove the `tinyforest` environment. Say "no" unless you have a previous installation and are having trouble. It will create a clean install of the VE.

-   Installs a number of Python packages to the `tinyforest` environment. These include

    -   `earthengine-api` which enables access to Google Earth Engine (GEE)

    -   `geemap` - a set of added value tools to extract and manipulate GEE data

    -   `osdatahub` - access to Ordnance Survey National Geographic Database data.

    -   `OSGridConverter` - converts lat-longs to UK Grid references

-   Imports the relevant modules for use in other packages

### Install and initialise

```{r setup, cache=FALSE}
if(!require("tinyForestR"))
#devtools::install_github("julianflowers/tinyForestR", force = FALSE)
if(!require("pacman"))install.packages("pacman")
library(pacman)
library(tinyForestR)

p_load(leaflet.extras2, tidyverse, mapview, sf, ggmap, lubridate, geojsonio)

theme_set(ggthemes::theme_economist())

initialise_tf()
```

## Load Tiny Forest data

The next step is to load Tiny Forest (TF) data. Because this only exists in a series of web pages the `get_tf_data` function identifies the relevant pages and iterates over them to extract name, id, location, area, planters, and types of tree planted (as a list column), for those TFs planted at the time of extraction. It does include TFs planted outside the UK. The function takes about 30 seconds to iterate over all the relevant pages.

```{r, tf_data}

tf <- tinyForestR::df

tf_df <- tf |>
  unnest("trees") |>
  mutate(year = year(date), 
         month = month(date)) 



```

Once the data is loaded we can save it as a csv file and get some high level information on planting, timings, size and so on.

As of `r today()` there are `r length(unique(tf_df$tf_id))` planted TFs.

```{r tf-stats, fig.height=6, fig.width=8}

needs(patchwork)

## annual planting

tf_year <- tf_df |>
  select(-trees) |>
  distinct() |>
  count(year) |>
  ggplot(aes(year, n)) +
  geom_col() +
  labs(title = "TFs planted per year")

## area distribution

tf_area <- tf_df |>
  select(-trees) |>
  distinct() |>
  ggplot(aes(factor(year), area)) +
  geom_boxplot() +
  labs(title = "TF area year")

tf_trees <- tf_df |>
  group_by(tf_id) |>
  summarise(n_trees = n()) |>
  ggplot() +
  geom_histogram(aes(n_trees)) +
  labs(title = "Distribution of tree species", 
       x = "Number of tree species")

tf_year + tf_area + tf_trees


```

## Locations

It is now straightforward to map locations of TFs using `sf` and `mapview`.

```{r map-location}

library(mapview)

tf_df |>
  select(-trees) |>
  distinct() |>
  mutate(year = factor(year)) |>
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  mapview::mapview(zcol = "year")

```

## Tree species

We can also look at planting frequency for different tree species.

```{r fig.height=6}

tf_df |>
  ungroup() |>
  unnest("trees") |>
  count(trees, sort = TRUE) |>
  top_n(25) |>
  ggplot() +
  geom_col(aes(n, reorder(trees, n))) +
  labs(y = "", 
       title = "25 most commonly planted tree species", 
       x = "Number of trees") +
 # ggthemes::theme_base() +
  theme(plot.title.position = "plot")

tf_df |>
  ungroup() |>
  unnest("trees") |>
  count(tf_id, trees, sort = TRUE) |>
  pivot_wider(names_from = trees, values_from = n, values_fill = 0) 
  
  
  

```

## Biodiversity data

```{r get-lat-lon}

tf1 <- tf_df |>
 select(lat, lon, tf_id) |>
  distinct()

i <- 1

lat <- tf1$lat[i]
lon <- tf1$lon[i]
```

The `get_nbn_buffer` downloads occurrence data from the NBN Atlas in a set buffer around a given longitude and latitude. For example we can download 10000 records around lat=`r lat`, lon=`r lon` (Witney TF).[^1]

```{r biodiversity-data}

safe_buff <- safely(tinyForestR::get_nbn_buffer)

nbn_data <- safe_buff(lon, lat, n = 10000)

nbn_data$result |>
  head()

```

## Plant diversity using BSBI data

I have also included functions to extract data for the 2020 Botanic Society of Britain and Ireland survey. This is publicly available from for UK National Grid 1k hectads. This requires conversion of lat-longs to UK grids.

```{r os-to-grid}

grid_ref <- tinyForestR::os_lat_lon_to_grid(lat = lat, lon = lon)

grid_ref$grid

```

```{r bsbi-data}

bsbi_data <- tinyForestR::get_bsbi_data(grid_ref = grid_ref$grid)

bsbi_data |>
  enframe() |>
  unnest("value") |>
  unnest("value") |>
  # slice(-c(168:173)) |>
  mutate(year = str_extract(value, "20\\d{2}"),
         value = str_remove(value, year),
         count = parse_number(value),
         value = str_remove(value, as.character(count)), 
         value = str_remove(value, "\\d{1,}"), 
         grid = grid_ref$grid, 
         tf_id = tf_df$tf_id[i]) |>
  arrange(value)  |>
  drop_na()
```

## Rapid calculation of biodiversity metrics

The `calc_bd_metrics` function takes an output from `get_nbn_buffer` or `get_bsbi_data`, converts the data from long to wide format, creates a species matrix for a specified class (for `get_nbn_buffer` data), and outputs a list containing:

-   diversity metrics (Simpson)
-   species richness by month (plotted)
-   Bray-Curtis matrix
-   Raup matrix

```{r biodiversity-metrics}

metrics <- calc_bd_metrics(df = nbn_data$result, class = "Aves")


metrics$metrics
```

```{r}
metrics$plot +
  labs(title = "Monthly species richness for ",
         subtitle = paste("Aves", tf1$tf_id[1]), 
       y = "Richness", 
       x = "Month") +
  theme(plot.title.position = "plot")
```

## Vegatation indices

The package includes a `calc_ndvi_buff` function to enable the calculation of normalized vegetation index (NDVI) for the buffer area around a given point. It uses Sentinel2 surface reflectance satellite images which are available at 10m resolution and are regularly updated. The function extracts images via the Google Earth Engine API and requires registration and authentication prior to use (see...).

The function returns a list including, image dates, NDVI statistics for the image, an interactive map and a raster. Note it may take few minutes to run.

The code chunk below calculates the NDVI for each image containing the buffer around the Witney TF for 2019 and 2022 and maps them side-by-side. (Note, the function selects only those S2 images with cloud cover \< 10%).

```{r ndvi-images, eval=TRUE , cache=TRUE}

lon <- df$lon |> 
  unique()

lat <- df$lat |> 
  unique()

tf_id <- df$tf_id |> 
  unique()

plant_date <- df$date |>
  unique()

i <- 10

test <- calc_ndvi_buff(lon = lon[i], lat = lat[i], dist1 = 1000, dist2 = 100, start_date = as.character(plant_date[2]), end_date = "2023-06-01", cloud_cover = 10, tf_id = tf_id[i])


test1 <- calc_ndvi_buff(lon = lon[i],  lat =  lat[i], dist1 = 1000, dist2 = 100, start_date = "2017-01-01", end_date = as.character(plant_date[2]), cloud_cover = 10, tf_id = tf_id[i])


test1$map | test$map


```

## 

```{r}

pre <- test1$point_summary |>
  mutate(period = "a-pre", 
         loc = "point") |>
  mutate_at(.vars = 3:5, scale)

post <- test$point_summary |>
  mutate(period = "b-post", 
         loc = "point") |>
  mutate_at(.vars = 3:5, scale)

pre_buff <- test1$buffer_summary |>
  mutate(period = "a-pre", 
         loc = "buffer") |>
  mutate_at(.vars = 3:5, scale)

post_buff <- test$buffer_summary |>
  mutate(period = "b-post", 
         loc = "buffer") |>
  mutate_at(.vars = 3:5, scale)

points <- bind_rows(pre, post)
buffers <- bind_rows(pre_buff, post_buff)

overall <- bind_rows(points, buffers)

overall |>  
  ggplot() +
  geom_boxplot(aes(loc, ndvi_mean, fill = period)) 

overall |>
  ggplot() +
  geom_col(aes(date, ndvi_median, fill = loc), position = "dodge") +
  geom_smooth(aes(date, ndvi_median, color = loc), method = "loess", span = 0.5) +
  geom_vline(xintercept = as.Date(plant_date[i]), lty = "dotted") +
  geom_hline(yintercept = 0) +
  facet_wrap(~ loc) +
  labs(title = "Trend in scaled NDVI: 100m and 1000m from TF centroid",
       subtitle = paste0("Tiny Forest ", tf_id[i], "; planted: ", plant_date[i]), 
       x = "")
```

## Covariates

TF specific:

-   Size

-   Tree diversity

-   Age

-   Urban / rural

Area specific

-   Green infrastructure (conducive to biodiversity)[@WOS:000882918000004; @WOS:000327304100013; @WOS:000572856700001; @WOS:000374275700069; @implemen; @jerome2018; @Van2020; @Felappi2020; @Yin2022; @Anderson2022][@burgess2015; @strategi; @jones2022]

I have included a snapshotwith environmental variables for the TF dataset. This includes the age of the TF, rural-urban classification of the lower super output area containing the TF centroid, the area (m^2^) of public parks and allotments in the buffer taken from OS Open Greenspace data, and the area (m^2^) of deciduous woodland, taken from Priority Habitat Inventory data.

### Typology of green infrastructure

[^1]: Note this may sometimes time out depending on traffic on the NBN Atlas webservice
